Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-19 draft.tex
changed abstract with help of Kai

%% indicate IEEE Member or Student Member in form indicated below
\author{TBA}
\authorfooter{
%% insert punctuation at end of each item
% \item
%  Otto-v.-Guericke-University Magdeburg
}

%other entries to be set up for journal
\shortauthortitle{Klemm \MakeLowercase{\textit{et al.}}: Interactive Visual Analytics of Image-Centric Cohort Study Data}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
%Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{interactions}).
%
Modern cohort studies also incorporate medical image data.
%
Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
%As extension to the epidemiological workflow it aims to unveil interactions based on all reliable shape information rather than abstracting it to a few key figures.
%Iterative analysis between ... allows for extracting interactions which without the visiual analysis driven approach maybe would be overlooked. %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen 
%Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
%Iterative analysis of image-based epidemiological datasets aims to provide insight into interactions between organ shape and % in order to provide insights into \emph{interactions} by incorporating additional information
%It extends the epidemiological workflow with an iterative Analysis process and 
It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Interactive Visual Analytics, Epidemiology}

@@

\subsection{Data Preprocessing}
To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
%
Due to the different acquisition modalities there have to be different techniques incorporated.
%
Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS} or \texttt{Stata} which have a proprietary data format with limited export capabilities.
%
The best solution for us was to simply export the data in the respective tool to a character separated text file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
%
In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
%
Exporting the data dictionary, which stores information about each manifestation of a feature is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
%
The reasons for missing data have a wide range from ethical to medical and personal issues.
%
Therefore, these are also included as error codes which have to marked as such in the data dictionary.
\\\\
Processing the image data associated to each subject consists for the most part of information extraction about a structure of interest.
%
This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as Graph-Cuts TODO: REFERENCE) or by a (semi)-automatic detection, registration and segmentation.
%
Algorithms, applied to the data, do not only have to deal with a large inter-subject variability of the structure of interest but also need to be reproducible \cite{Preim2014}.
%
Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
%
If a segmentation yields only binary masks separating the structures, algorithms such as Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
%
Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
%
Comparison based on grey values is usually carried out to to measuring the quantity of fat, water, and--application specific--iron content (liver) or distribution of grey and white brain tissue.

@@

\section{Application}
\subsection{The Spine Dataset}
- Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
- Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
- Input of Epidemiologists goes here!\\
\\\\
\textbf{From VMV'13 Paper}\\

@@

\section{Summary and Conclusion}

>>>>

%% indicate IEEE Member or Student Member in form indicated below
\author{--blind--}
\authorfooter{
%% insert punctuation at end of each item
% \item
%  Otto-v.-Guericke-University Magdeburg
}

%other entries to be set up for journal
\shortauthortitle{Klemm \MakeLowercase{\textit{et al.}}: Interactive Visual Analysis of Image-Centric Cohort Study Data}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
\abstract{%
%(The problem). 
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{interactions}).
%
Modern cohort studies also incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D-shape renderings with epidemiological data representations such as pivot tables and key figures as odds ratios.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypothesis about the aging process of the lumbar spine.
%
%(Results). 
It shows to be a more flexible comparison between image-and non-image data.
%
%(Implications).
The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.

%
%Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
%Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
} % end of abstract

% \abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
% %
% Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
% %
% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{interactions}).
% %
% Modern cohort studies also incorporate medical image data.
% %
% Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
% %
% The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
% \\\\
% We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
% %
% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
% %As extension to the epidemiological workflow it aims to unveil interactions based on all reliable shape information rather than abstracting it to a few key figures.
% %Iterative analysis between ... allows for extracting interactions which without the visiual analysis driven approach maybe would be overlooked. %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen 
% %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
% %Iterative analysis of image-based epidemiological datasets aims to provide insight into interactions between organ shape and % in order to provide insights into \emph{interactions} by incorporating additional information
% %It extends the epidemiological workflow with an iterative Analysis process and 
% It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
% %
% Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
% %
% Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
% %
% Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
% } % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Interactive Visual Analysis, Epidemiology}

@@

\subsection{Data Preprocessing}
To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
%
Due to the different acquisition modalities there have to be different techniques incorporated.
%
Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS} or \texttt{Stata} which have a proprietary data format with limited export capabilities.
%
The best solution for us was to simply export the data in the respective tool to a character separated text file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
%
In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
%
Exporting the data dictionary, which stores information about each manifestation of a feature is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
%
The reasons for missing data have a wide range from ethical to medical and personal issues.
%
Therefore, these are also included as error codes which have to marked as such in the data dictionary.
\\\\
Processing the image data associated to each subject consists for the most part of information extraction about a structure of interest.
%
This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as graph cuts \cite{GraphCut}) or by a (semi)-automatic detection, registration and segmentation.
%
Algorithms, applied to the data, do not only have to deal with a large inter-subject variability of the structure of interest but also need to be reproducible \cite{Preim2014}.
%
Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
%
If a segmentation yields only binary masks separating the structures, algorithms such as Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
%
Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
%
Comparison based on grey values is usually carried out to to measuring the quantity of fat, water, and--application specific--iron content (liver) or distribution of grey and white brain tissue.

@@

\section{Application}
- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way
\subsection{The Spine Dataset}
- Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
- Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
- Input of Epidemiologists goes here!\\
\\\\
\textbf{From VMV'13 Paper}\\

@@

\section{Summary and Conclusion}
- Future Work: Matrix-View of differences - more intuitive\\
- UI more flexible by hiding UI-panes
